---
title: "Part I: Explore. Data visualisation"
output: html_notebook
---

library(tidyverse)
library(ggplot2)
options(tibble.print_max = 20, tibble.print_min = 8)

data("mpg")


### 3.2.4 Exercises

1. Run ggplot(data = mpg). What do you see?

ggplot(data = mpg) creates an empty graph. The ggplot() creates a coordinate system to which the layers are added and takes the dataset as its first argument.

2. How many rows are in mpg? How many columns?
```{r}
dim(mpg)
```
There are 234 rows and 11 columns in mpg.

3. What does the drv variable describe? Read the help for ?mpg to find out.
```{r}
?mpg
```
drv describes the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd.

4. Make a scatterplot of hwy vs cyl.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(hwy, cyl))
```


5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?
  
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(class, drv))
```

Both drv and class are categorical variables, while scatterplots are more suitable for representing continuous values. In the plot above we cannot see how many observations belong to a single point.


### 3.3.1 Exercises

1. What’s gone wrong with this code? Why are the points not blue?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

The argument colour = "blue" should be defined outside the aes():
The aesthetic here is a mapping between a variable and a value and colour = "blue" put inside it is interpreted as another categorical variable with a single value.

2. Which variables in mpg are categorical? Which variables are continuous?

```{r}
glimpse(mpg)
```
Categorical: manufacturer, model, trans, drv, fl, class.
Continuous: displ, year, cyl, cty, hwy


3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?
```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour = year)) +
  geom_point()
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = year)) +
  geom_point()
```
```{r}
ggplot(mpg, aes(x = displ, y = hwy, shape = year)) +
  geom_point()
```
Error: A continuous variable can not be mapped to shape


4. What happens if you map the same variable to multiple aesthetics?
```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour = year, size = year)) +
  geom_point()
```
Mapping a single variable to multiple aesthetics makes the chart illegible and is redundant.

5. What does the stroke aesthetic do? What shapes does it work with? 

The stroke aesthetic modifies the width of the border of a shapes (for shapes 21-25 the inside and outside can be coloured separately).

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.


```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour = displ < 5)) +
  geom_point()

```

Mapping to expressions adds a temporary variable to the data - for the result of "displ < 5" it is a logical variable which takes values of TRUE or FALSE.

### 3.5.1 Exercises

1. What happens if you facet on a continuous variable?

The continuous variable would be converted to a categorical type and facets for each distinct value would be plotted.

2. What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?

The empty cells in this plot mean that have no observations.

3. What plots does the following code make? What does . do?

The "." ignores that dimension when faceting: drv ~ . facet by values of drv on the y-axis, . ~ cyl will facet by values of cyl on the x-axis.

4. What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

Pros for facets:
- increase the ability to encode more distinct categories
- easier to compare the shape of the relationship between the x and y variables across categories
- it is difficult to handle overlapping points with color.

Cons for facets:
- difficult to directly compare values of observations between categories since the points are spread across multiple plots.


5. Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?

nrow, ncol	- number of rows and columns. 
In facet_grid() the number of rows and columns are determined by the number of unique values of the variables specified in the function.

6. When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?

More unique levels in the columns will leave more space for columns if the plot is laid out horizontally.





### 3.6.1 Exercises

1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
geom_line() - line chart
geom_boxplot() - boxplot
geom_histogram() - histogram
geom_area() - area chart

3. What does show.legend = FALSE do? 
It hides the legend box.

5. What does the se argument to geom_smooth() do?
It adds standard error bands to the lines.

6. Will these two graphs look different? Why/why not?

No, because the same data and mappings will be used by geom_point() and geom_smooth() The geom_smooth() will inherit those arguments from the ggplot() object, so there is no need to specify mappings again.

7. Recreate the R code necessary to generate the following graphs.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(mapping = aes(group = drv), se = FALSE)

ggplot(data = mpg, aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) +
  geom_smooth(se = FALSE)

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) 

```



### 3.7.1 Exercises

1. What is the default geom associated with stat_summary()? How could you rewrite the previous plot to use that geom function instead of the stat function?

Default: geom_pointrange()
For example by changing argument identity() to stat = "summary" (stat_summary()).

2. What does geom_col() do? How is it different to geom_bar()?

geom_col():
- default stat is stat_identity(), which leaves the data as is 
- data passed as argumnet must contain x values and y values which represent the bar height

geom_bar()
- default stat is stat_bin() 
- only x variable must be passed as an argument 
- input data are preprocessed by counting the number of observations for each value of x. The y aesthetic uses the values of these counts.


4. What variables does stat_smooth() compute? What parameters control its behaviour?

Computed variables:
y - predicted value
ymin - lower pointwise confidence interval around the mean
ymax - upper pointwise confidence interval around the mean
se - standard error

5. In our proportion bar chart, we need to set group = 1. Why? In other words what is the problem with these two graphs?

If group = 1 is not included then proportions are calculated with groups and all of the bars have equal height (1 --> 100%).





### 3.8.1 Exercises

1. What is the problem with this plot? How could you improve it?

The problem is in overplotting and it can be fixed by adding jitter.
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point(position = "jitter")
```

2. What parameters to geom_jitter() control the amount of jittering?
width	- Amount of vertical and horizontal jitter.
height - Amount of vertical and horizontal jitter. 

3. Compare and contrast geom_jitter() with geom_count().

geom_jitter():
- pros - reduces overplotting while adding random variation to the locations of the points on the graph
- cons - it slightly changes the x and y values of the points.

geom_count():
- pros - does not change x and y coordinates of the points
- cons - if the points are close together and counts are large, the size of some points can overplot the other once

4. What’s the default position adjustment for geom_boxplot()? Create a visualisation of the mpg dataset that demonstrates it.

Default: dodge2

```{r}
ggplot(data = mpg, aes(x = drv, y = hwy, fill = class)) +
  geom_violin(position = "identity")
```




### 3.9.1 Exercises

1. Turn a stacked bar chart into a pie chart using coord_polar().
```{r}
ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar()
```



2. What does labs() do? Read the documentation.
labs() adds x- and y-axis titles, plot titles, subtitles and a caption to the plot.

3. What’s the difference between coord_quickmap() and coord_map()?

coord_map():
- projects the three-dimensional Earth onto a two-dimensional plane
- uses the Mercator projection by default

coord_quickmap():
- uses an approximate but faster map projection
- ignores the curvature of Earth and adjusts the map for the latitude/longitude ratio. 

